Generalized spaces for constructive algebra

### 2020-04-04 paper-herrsching2019/main.tex
Gelfand duality and several smaller items

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\ab}{\mathrm{ab}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\Gal}{\mathrm{Gal}}
\newcommand{\Rad}{\mathrm{Rad}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Pt}{\mathrm{Pt}}
\newcommand{\Loc}{\mathrm{Loc}}
\newcommand{\Top}{\mathrm{Top}}
\newcommand{\BPIT}{\textsc{bpit}\xspace}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\negg}{\neg\!\!\!\neg}
\newcommand{\bott}{\bot\!\!\!\!\bot}

@@

% XXX \cite{belanger-marquis:pointless} and Johnstone, vickers:locales-toposes

@@

\begin{quote}
\textbf{Definition.} A \emph{metric space} consists of a set~$X$ of points
together with a metric~$d : X \times X \to \RR_{\geq 0}$ satisfying the metric
axioms.

@@

By relinquishing points, locales provide a more flexible notion of space.
Section~\ref{sect:examples-no-points} substantiates this observation
with several examples of nontrivial locales without any points. Constructive
mathematics gives a further, orthogonal motivation to study locales: There are
situations in which the relevant spaces do have enough points, but only if one
subscribes to the axiom of choice or similar non-constructive principles. In
these situations, the pointfree approach allowed by locales help to give
constructive versions of classical results. We will further discuss this theme
in Section~\ref{sect:presenting-frames}.

@@

\paragraph{Compactness of the unit interval.} The unit interval, when realized
as a topological space, can fail to be be compact in constructive
mathematics. For instance, it fails in the Russian school because the Kleene
tree provides a computable open covering of~$[0,1]$ with no computable finite
subcovering (a self-contained exposition of this phenomenon is
Ref.~\refcite{bauer:kleene-tree}).

@@

\paragraph{Gelfand duality} XXX

@@

The starting point of the theory of locales is the following
observation. The set~$\O(X)$ of open subsets of a topological
space~$X$ forms a partially ordered set which has
\[ \begin{array}{@{}ccc@{}}
\text{arbitrary joins}
&\text{and}&
\text{finite meets}, \\
\bigvee && \wedge
\end{array} \]
and finite meets distribute over arbitrary joins:
\[ U \wedge \bigvee_i V_i = \bigvee_i (U \wedge V_i). \]

@@

Examples for locales include the following.
\begin{enumerate}
\item Any topological space~$Y$ induces a locale~$L(Y)$ by
setting~$\O(L(Y)) \defeq \O(Y)$. A continuous map~$f : Y \to Y'$ of topological
spaces induces the frame homomorphism~$\O(Y') \to \O(Y),\,U \mapsto f^{-1}[U]$
in the other direction and hence a morphism~$L(Y) \to L(Y')$ of locales in the
same direction.
\item The \emph{one-point locale}~$\pt$ is the locale induced
by the one-point topological space~$\{\star\}$. Its frame of opens is
the powerset of~$\{\star\}$, also known as the set~$\Omega$ of truth
values. Its least element is~$\bot = \emptyset$ and its largest element
is~$\top = \{\star\}$, and potentially not all elements of~$\Omega$ are equal
to one of these two.
\item The locale of surjections~$\NN \twoheadrightarrow \RR$ and the localic
version of the reals of Section~\ref{sect:examples-no-points} are best
constructed as \emph{classifying locales}, a notion to be introduced below.
\end{enumerate}

@@

\begin{remark}The set of points of a locale~$X$ can be made into a topological space, giving
rise to a functor~$\Pt : \Loc \to \Top$. This functor is right adjoint
to the functor~$L : \Top \to \Loc$. A locale~$X$ is spatial iff the canonical
morphism~$L(\Pt(X)) \to X$ is an isomorphism, and a topological space~$Y$
is \emph{sober} iff the canonical morphism~$Y \to \Pt(L(Y))$ is a
homeomorphism. The space~$\Pt(L(Y))$ is the \emph{sobrification} of~$Y$; for
instance, the sobrification of any inhabited indiscrete space is the one-point
space.\end{remark}

\begin{remark}\label{rem:sober-embed}
Since continuous maps between spaces induce locale morphisms, we
have a canonical map~$\Hom_\Top(Y,Y') \to \Hom_\Loc(L(Y),L(Y'))$. If~$Y'$ is
sober, then this map is a bijection. In particular, the set-theoretic points
of~$Y'$ and the locale-theoretic points of~$L(Y')$ are in canonical
one-to-one correspondence.
\end{remark}

@@

@@

@@

\begin{exercise}{Coproducts of locales}%
\begin{enumerate}
\item Show that the category of frames has small products.
\item Deduce that the category of locales has small coproducts.
\item Let~$M$ be a set and endow it with the discrete topology. Show
that~$L(M)$ is the coproduct~$\coprod_{x \in M} \pt$.
\end{enumerate}
The category of locales also has small products;
Exercise~\ref{ex:products-of-locales} is devoted to this fact.
\end{exercise}

\begin{exercise}{Local locales}%
A locale~$X$ is \emph{local} iff for any open covering~$\top = \bigvee_i U_i$,
there is an index~$i$ such that~$\top = U_i$. (For instance, the spectrum of a
ring~$A$ is local iff~$A$ is local.)
Show that any local locale has a canonical point, its \emph{focal point}.
\end{exercise}

@@

\paragraph{The geometric theory of Dedekind cuts.}\label{par:dedekind-cuts}
This theory has no sorts; no
function symbols; nullary relation symbols~$\alpha_x$ and~$\beta_x$ for each
rational number~$x$ (to be read as~``$x$ is contained in the lower respectively
the upper cut''); and the following axioms, spelling out that the cut is bounded,
rounded, open and located:
\begin{align*}
  \top &\vdash \textstyle\bigvee_{x \in \QQ} \alpha_x \\
  \top &\vdash \textstyle\bigvee_{x \in \QQ} \beta_x \\
  \alpha_y &\vdash \alpha_x & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
  \beta_x &\vdash \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
  \alpha_x &\vdash \textstyle\bigvee_{y > x} \alpha_y & \text{(for each~$x \in \QQ$)} \\
  \beta_y &\vdash \textstyle\bigvee_{x < y} \beta_x & \text{(for each~$y \in \QQ$)} \\
  \alpha_x \wedge \beta_y &\vdash \bot & \text{(for each~$x,y \in \QQ$ such that~$x \geq y$)} \\
  \top &\vdash \alpha_x \vee \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)}
\end{align*}

@@

\begin{remark}\label{rem:traditional-spaces}
The traditional way of constructing the real line or the Zariski
spectrum as a topological space proceeds in three steps: (1) Write down the
axioms (of Dedekind cuts respectively prime filters). (2) Using the powerset
axiom and separation, construct the set of Dedekind reals respectively the set
of prime filters. (3) Devise a useful topology on the resulting set.

@@

\subsection{Sheaf semantics}

@@

@@

The story of generic models is at the same time tautologous and deep, and these
short paragraphs do not do justice to this intriguing aspect of the interplay
between topology (in the form of locales) and logic (in the form of
propositional geometric theories). We strongly
recommend the accessible account by Steve Vickers~\ref{vickers:continuity} for a fuller
picture.

@@

{\scriptsize\emph{Note.} In the universe of sheaves over this locale, the set
of unit interval Dedekind real numbers fails to be (Heine--Borel) compact and
the localic real line fails to be
spatial~\cite[Example~D4.7.13]{johnstone:elephant}.\par}
\item An open~$U$ of a locale~$X$ is \emph{complemented} iff there is an
open~$V$ such that~$U \vee V = \top$ and~$U \wedge V = \bot$. A locale is
\emph{zero-dimensional} iff any open is a join of complemented opens. Assuming
that any two reals are equal or not, show that the locale of surjections~$\NN
\to \RR$ is zero-dimensional.
\end{enumerate}
\end{exercise}

@@

\begin{exercise}{Product of locales}\label{ex:products-of-locales}%
Let~$X$ and~$X'$ be locales, classifying some propositional geometric
theory~$\TT$ respectively~$\TT'$. Let~$\TT + \TT'$ be the combined theory
(whose set of sorts is the disjoint union of the sets of sorts of~$\TT$
and~$\TT'$, and so on). Show that~$L(\TT + \TT')$ is the product of~$X$
and~$X'$ in the category of locales.
\end{exercise}

\begin{exercise}{Coproduct of locales, revisited}%
Let~$X$ and~$X'$ be locales, classifying some propositional geometric
theory~$\TT$ respectively~$\TT'$. Find a theory in terms of~$\TT$
and~$\TT'$ which is classified by~$X \amalg X'$.
\end{exercise}

@@

\item \emph{Deligne's completeness theorem} states that coherent locales (in
fact, even coherent \emph{toposes}) are spatial. Show that Deligne's
completeness theorem implies that propositional coherent theories
without any set-based models are inconsistent.
\end{enumerate}
\end{exercise}

@@

\begin{exercise}{On the failure of the fundamental theorem of algebra}%
Let~$F$ be the sheaf of continuous complex-valued functions on~$\CC$. Give an
example of a function~$f : \CC \to \CC$ such that it is not the case that
\[ \Sh(X) \models \exists z\?F\_ z^2 = f. \]
{\scriptsize\emph{Note.} The sheaf~$F$ can be obtained by constructing, in the
universe of sheaves on~$\CC$, the set of pairs of Dedekind reals. Hence the
example demonstrates that the usual formulation of the fundamental theorem of
algebra does not admit a constructive proof. (However, there are replacements
available~\cite{richman:fta}, and the fundamental theorem does apply to the
Cauchy complex numbers~\cite{ruitenburg:roots} and under weak choice
principles~\cite{bridges-richman-schuster:wcc}).\par}
\end{exercise}

@@

Such an understanding is imparted by the following theorem. This theorem is the
central workhorse of our approach.

@@

\begin{exercise}{Free fields}
\begin{enumerate}
\item A \emph{free field} over a set~$M$ would be
a field~$K$ together with a map~$M \to K$ such that any map~$M \to L$ into a
field~$L$ factors
over the given map~$M \to K$ by a unique field homomorphism~$K \to L$. Prove
that free fields exist for no set~$M$.
\item Let~$k$ be a field. Devise a hypothetical notion of
\emph{free~$k$-fields} (where a~$k$-field is a field together with a
homomorphism from~$k$) and show that there are no free~$k$-fields.
\end{enumerate}
\end{exercise}

\begin{exercise}{Nontriviality of~$A^\sim$}%
Let~$A$ be a ring. In Proposition~\ref{prop:immediate-consequences}, we gave an
abstract proof that~$A^\sim$ is local and hence nontrivial in the sense that~$1
\neq 0$ (even if~$A$ itself happens to be trivial). Give a new proof of this
fact, by verifying directly~$1 \models \neg(1 =_{A^\sim} 0)$ using the
semantics given in Table~\ref{table:algebraic-kripke-joyal}.
\end{exercise}

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\ab}{\mathrm{ab}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\Max}{\mathrm{Max}}
\newcommand{\Gal}{\mathrm{Gal}}
\newcommand{\Rad}{\mathrm{Rad}}
\newcommand{\Idl}{\mathrm{Idl}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Pt}{\mathrm{Pt}}
\newcommand{\Loc}{\mathrm{Loc}}
\newcommand{\Top}{\mathrm{Top}}
\newcommand{\BPIT}{\textsc{bpit}\xspace}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\negg}{\neg\!\!\!\neg}
\newcommand{\bott}{\bot\!\!\!\!\bot}

@@

% XXX \cite{belanger-marquis:pointless} and Johnstone, vickers:locales-toposes,
% picado-pultr:frames-and-locales

@@

% NOTE TO EDITORS AND LAYOUTERS: It is on purpose that these two definitions
% are not set using the usual theorem environment. This is because we are only
% citing these definitions in order to reflect on them. We are not actually
% introducing these two notions here for further use.
\begin{quote}
\textbf{Definition.} A \emph{metric space} consists of a set~$X$ of points
together with a metric~$d : X \times X \to \RR_{\geq 0}$ satisfying the metric
axioms.

@@

By relinquishing points, locales provide a more flexible notion of space.
Section~\ref{sect:examples-no-points} substantiates this observation
with several examples of nontrivial locales without any points. Constructive
mathematics gives a further, orthogonal motivation to study locales: There are
situations in which the relevant spaces do have enough points, but only if one
subscribes to the axiom of choice or similar non-constructive principles. In
these situations, the pointfree approach allowed by locales helps to give
constructive versions of classical results. We will further discuss this theme
in Section~\ref{sect:presenting-frames}.

@@

\paragraph{Compactness of the unit interval.} The unit interval, when realized
as a topological space, can fail to be be compact in constructive
mathematics. For instance, it fails in the Russian school because the Kleene
tree provides a computable open covering of~$[0,1]$ with no computable finite
subcovering (a self-contained exposition of this phenomenon is contained in
Ref.~\refcite{bauer:kleene-tree}).

@@

\paragraph{Gelfand duality.} The celebrated Gelfand correspondence states that
the mapping
\[ X \longmapsto \Hom(X, \CC) = \{ f : X \to \CC \,|\, \text{$f$ is continuous} \} \]
defines a contravariant equivalence between compact Hausdorff spaces and
commutative unital~$C^\star$-algebras. In the other direction,
a~$C^\star$-algebra~$A$ is mapped to its \emph{maximal spectrum}~$\Max(A)$, the
topological space of multiplicative linear functionals~$A \to \CC$.

Some parts of the proof require the axiom of choice, (only) for fabricating
appropriate elements of the maximal spectrum. But as with the fundamental
theorem of infinite Galois theory, the Gelfand correspondence has a
constructive core, and a pointfree reformulation unearths this constructive
content: There is a contravariant equivalence between compact completely
regular locales and commutative unital~$C^\star$-algebras.
% XXX Barr

@@

The starting point of the theory of locales is the following
observation. The set~$\O(X)$ of open subsets of a topological
space~$X$ forms a partially ordered set which has
\[ \begin{array}{@{}ccc@{}}
\text{arbitrary joins (suprema)}
&\quad\text{and}\quad&
\text{finite meets (infima)}, \\
\bigvee && \wedge
\end{array} \]
and finite meets distribute over arbitrary joins:
\[ U \wedge \bigvee_i V_i = \bigvee_i (U \wedge V_i). \]

@@

Examples for locales include the following.
\begin{enumerate}
\item Any topological space~$Y$ induces a locale~$L(Y)$ by
setting~$\O(L(Y)) \defeq \O(Y)$. A continuous map~$f : Y \to Y'$ of topological
spaces induces the frame homomorphism~$\O(Y') \to \O(Y),\,U \mapsto f^{-1}[U]$
in the other direction and hence a morphism~$L(Y) \to L(Y')$ of locales in the
same direction.
\item The \emph{one-point locale}~$\pt$ is the locale induced
by the one-point topological space~$\{\star\}$. Its frame of opens is
the powerset of~$\{\star\}$, also known as the set~$\Omega$ of truth
values. Its least element is~$\bot = \emptyset$ and its largest element
is~$\top = \{\star\}$, and potentially not all elements of~$\Omega$ are equal
to one of these two.
\item The locale of surjections~$\NN \twoheadrightarrow \RR$ and the localic
version of the reals of Section~\ref{sect:examples-no-points} are best
constructed as \emph{classifying locales}, a notion to be introduced in
Section~\ref{sect:presenting-frames}.
\item The radical ideals of any ring~$A$ form a frame and hence give rise to a
locale. We will identify this locale in Theorem~\ref{thm:workhorse} to be the
\emph{spectrum} of~$A$ as studied in algebraic geometry.
\end{enumerate}

@@

\begin{remark}\label{rem:definition-sober}
The set of points of a locale~$X$ can be made into a topological space, giving
rise to a functor~$\Pt : \Loc \to \Top$. This functor is right adjoint
to the functor~$L : \Top \to \Loc$. A locale~$X$ is spatial iff the canonical
morphism~$L(\Pt(X)) \to X$ is an isomorphism, and a topological space~$Y$
is \emph{sober} iff the canonical morphism~$Y \to \Pt(L(Y))$ is a
homeomorphism. The space~$\Pt(L(Y))$ is the \emph{sobrification} of~$Y$; for
instance, the sobrification of any inhabited indiscrete space is the one-point
space.

The adjunction~$L \dashv \Pt$ restricts to an equivalence between sober
topological spaces and spatial locales. Assuming the law of excluded middle,
both Hausdorff spaces and the schemes from algebraic geometry are sober
(Exercise~\ref{ex:hausdorff-sober}), hence the spaces of a wide
range of the mathematical landscape can be studied as locales.\end{remark}

\begin{remark}\label{rem:sober-embed}
Since continuous maps between spaces induce locale morphisms, we
have a canonical map~$\Hom_\Top(Y,Y') \to \Hom_\Loc(L(Y),L(Y'))$. If~$Y'$ is
sober, then this map is a bijection. In particular, in this case the set-theoretic points
of~$Y'$ and the locale-theoretic points of~$L(Y')$ are in canonical
one-to-one correspondence.
\end{remark}

@@

\begin{exercise}{Hausdorff spaces are sober}%
\label{ex:hausdorff-sober}%
A \emph{meet-irreducible open subset} of a topological space~$Y$ is an
open subset~$W \subseteq Y$ such that~$(U \cap V \subseteq W) \Rightarrow (U
\subseteq W) \vee (V \subseteq W)$ for all open subsets~$U,V \subseteq Y$ and
such that~$Y \not\subseteq W$.
\begin{enumerate}
\item Assuming the law of excluded middle, show that the completely prime
filters of a topological space are in canonical one-to-one correspondence with
the meet-irreducible open subsets.

{\scriptsize\emph{Hint.} Given a completely prime filter~$\fff \subseteq
\O(Y)$, the set~$\bigcup\{ U \in \O(Y) \,|\, U \not\in \fff \}$ is a
meet-irreducible open subset.\par}
\item Assuming the law of excluded middle, show that a topological space is
sober in the sense of Remark~\ref{rem:definition-sober} if and only if for
every meet-irreducible open subset~$W$ there is a unique point~$y \in Y$ such
that~$W = Y \setminus \overline{\{y\}}$.
(This is a classical definition of sobriety.)

\item Assuming the law of excluded middle, show that Hausdorff topological
spaces validate the condition stated in part~(b) and hence are sober.
\end{enumerate}
\end{exercise}

@@

\begin{exercise}{On the covariant approach}%
Let~$A$ be a frame.
\begin{enumerate}
\item Show that~$A$ contains arbitrary (set-indexed) meets (though they are not
required to be preserved by frame homomorphisms), by the impredicative
construction
\[ \bigwedge_{i \in I} U_i = \bigvee\{ \text{$V \preceq U_i$ for all~$i \in I$}
\}. \]
\item What are the arbitrary meets in the frame of open subsets of a
topological space?
\item Show that for any frame homomorphism~$\alpha : A \to A'$ there exists a
monotone map~$\beta : A' \to A$ such that~\emph{$\beta$ is right adjoint
to~$\alpha$}, that is such that
\[ \alpha(U) \preceq V \quad\text{iff}\quad U \preceq \beta(V) \]
for all~$U \in A$ and~$V \in A'$.
\end{enumerate}
{\scriptsize\emph{Note.} Based on the observation in part~(c), a
\emph{covariant approach} to locales can be developed, in which locale maps~$X
\to X'$ are defined to be certain kinds of maps~$\O(X) \to
\O(X')$~\cite{picado-pultr:covariant}.\par}
\end{exercise}

@@

\begin{exercise}{The frame of radical ideals}%
\label{ex:frame-of-radical-ideals}%
Let~$A$ be a ring. We order the set~$\Idl(A)$ of ideals of~$A$ by inclusion.
(See Section~\ref{sect:algebraic-preliminaries} for some algebraic
preliminaries.)
\begin{enumerate}
\item Verify that~$\Idl(A)$ has arbitrary meets and joins, given by
intersection and sum of ideals, but that it is in general not a frame.

{\scriptsize\emph{Hint.} The distributive law fails in the polynomial ring~$\QQ[X,Y]$.\par}
\item Show that the sub-poset~$\Rad(A)$ of radical ideals is a frame, with
the same finite meets but with joins given by the radical of the join computed
in~$\Idl(A)$.
\item\label{item:join-top} Let~$\varphi$ be a truth value. Let~$\{\top\,|\,\varphi\}$ be the
subsingleton subset of~$\Rad(A)$ which contains the top element~$\sqrt{(1)}$
iff~$\varphi$. Verify that
\[ \bigvee\{\top\,|\,\varphi\} = \{ f \in A \,|\, \text{$f$ is nilpotent
or~$\varphi$} \} \]
in~$\Rad(A)$.
\end{enumerate}
\end{exercise}

\begin{exercise}{Coproducts of locales}%
\begin{enumerate}
\item Show that the category of frames has small products.
\item Deduce that the category of locales has small coproducts.
\item Let~$M$ be a set and endow it with the discrete topology. Show
that~$L(M)$ is the coproduct~$\coprod_{x \in M} \pt$.
\end{enumerate}
{\scriptsize\emph{Note.} The category of locales also has small products;
Exercise~\ref{ex:products-of-locales} is devoted to this fact.\par}
\end{exercise}

\begin{exercise}{Local locales}%
\label{ex:local-locales}%
A locale~$X$ is \emph{local} iff for any open covering~$\top = \bigvee_i U_i$,
there is an index~$i$ such that~$\top = U_i$. (For instance, the spectrum of a
ring~$A$ is local iff~$A$ is local.)
Show that any local locale has a canonical point, its \emph{focal point}.
\end{exercise}

@@

\paragraph{The geometric theory of Dedekind cuts.}\label{par:dedekind-cuts}
This theory has no sorts; no
function symbols; nullary relation symbols~$\alpha_x$ and~$\beta_x$ for each
rational number~$x$ (to be read as~``$x$ is contained in the lower (upper)
cut''); and the following axioms, spelling out that the cut is bounded,
rounded, open and located:
\begin{align*}
  \top &\vdash \textstyle\bigvee_{x \in \QQ} \alpha_x \\
  \top &\vdash \textstyle\bigvee_{x \in \QQ} \beta_x \\
  \alpha_y &\vdash \alpha_x & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
  \beta_x &\vdash \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)} \\
  \alpha_x &\vdash \textstyle\bigvee_{y > x} \alpha_y & \text{(for each~$x \in \QQ$)} \\
  \beta_y &\vdash \textstyle\bigvee_{x < y} \beta_x & \text{(for each~$y \in \QQ$)} \\
  \alpha_x \wedge \beta_y &\vdash \bot & \text{(for each~$x,y \in \QQ$ such that~$x \geq y$)} \\
  \top &\vdash \alpha_x \vee \beta_y & \text{(for each~$x,y \in \QQ$ such that~$x < y$)}
\end{align*}

@@

\begin{remark}\label{rem:traditional-spaces}
The traditional way of constructing the real line or the Zariski
spectrum as a topological space proceeds in three steps: (1) Write down the
axioms (of Dedekind cuts or of prime filters, respectively). (2) Using the powerset
axiom and separation, construct the set of Dedekind reals or the set
of prime filters, respectively. (3) Devise a useful topology on the resulting set.

@@

\subsection{Sheaf semantics}
\label{sect:sheaf-semantics}

@@

% XXX A sheaf may well be nontrivial even though it does not have any global
% sections. This is for instance the case for the subsheaf of the sheaf of continuous
% complex-valued functions on~$\CC$, given by~$U \mapsto \{ f : U \to \CC \,|\,
% \text{$f$ is continuous and~$f(z)^2 = z$ for all~$z \in U$} \}$. This sheaf
% has lots of local sections (local square root maps), but no global one.
% This phenomenon is related to the failure of (a simple version of) the
% fundamental theorem of algebra (Exercise~\ref{ex:fta}).

@@

The story of generic models is at the same time tautologous and deep, and these
short paragraphs do not do justice to this intriguing aspect of the interplay
between topology (in the form of locales) and logic (in the form of
propositional geometric theories). We strongly
recommend the accessible account by Steve Vickers~\cite{vickers:continuity} for a fuller
picture.

@@

{\scriptsize\emph{Note.} In the universe of sheaves over this locale, the set
of unit interval Dedekind real numbers fails to be (Heine--Borel) compact and
the localic real line fails to be
spatial~\cite[Example~D4.7.13]{johnstone:elephant}.\par}
\item An open~$U$ of a locale~$X$ is \emph{complemented} iff there is an
open~$V$ such that~$U \vee V = \top$ and~$U \wedge V = \bot$. A locale is
\emph{zero-dimensional} iff any open is a join of complemented opens. Assuming
that any two reals are equal or not, show that the locale of surjections~$\NN
\twoheadrightarrow \RR$ is zero-dimensional.
\end{enumerate}
\end{exercise}

@@

\begin{exercise}{The product of locales}\label{ex:products-of-locales}%
Let~$X$ and~$X'$ be locales, classifying propositional geometric
theories~$\TT$ and~$\TT'$ respectively. Let~$\TT + \TT'$ be the combined theory
(whose set of sorts is the disjoint union of the sets of sorts of~$\TT$
and~$\TT'$, and so on). Show that~$L(\TT + \TT')$ is the product of~$X$
and~$X'$ in the category of locales.
\end{exercise}

\begin{exercise}{The coproduct of locales, revisited}%
Let~$X$ and~$X'$ be locales, classifying some propositional geometric
theories~$\TT$ and~$\TT'$ respectively. Find a theory in terms of~$\TT$
and~$\TT'$ which is classified by~$X \amalg X'$.
\end{exercise}

@@

\item \emph{Deligne's completeness theorem} states that coherent locales (and
even coherent \emph{toposes}) are spatial. Show that Deligne's
result implies that propositional coherent theories
without any set-based models are inconsistent.
\end{enumerate}
\end{exercise}

@@

\begin{exercise}{On the failure of the fundamental theorem of algebra}%
\label{ex:fta}%
Let~$F$ be the sheaf of continuous complex-valued functions on~$\CC$. Give an
example of a function~$f : \CC \to \CC$ such that it is not the case that
\[ \Sh(X) \models \exists z\?F\_ z^2 = f. \]
{\scriptsize\emph{Note.} The sheaf~$F$ can be obtained by constructing, in the
universe of sheaves on~$\CC$, the set of pairs of Dedekind reals. Hence the
example demonstrates that the usual formulation of the fundamental theorem of
algebra does not admit a constructive proof. (However, there are replacements
available~\cite{richman:fta}, and the fundamental theorem does apply to the
Cauchy complex numbers~\cite{ruitenburg:roots} and under weak choice
principles~\cite{bridges-richman-schuster:wcc}).\par}
\end{exercise}

@@

Such an understanding is imparted by the following theorem. It is the
central workhorse of our approach.

@@

\begin{exercise}{Free fields}
\begin{enumerate}
\item A \emph{free field} over a set~$M$ would be
a field~$K$ together with a map~$M \to K$ such that any map~$M \to L$ into a
field~$L$ factors
over the given map~$M \to K$ by a unique field homomorphism~$K \to L$. Prove
that free fields exist for no set~$M$.
\item Let~$k$ be a field. Devise a hypothetical notion of
\emph{free~$k$-fields} (where a~$k$-field is a field together with a
homomorphism from~$k$) and show that the only set~$M$ such that there is a
free~$k$-field over~$M$ is the empty set.
\end{enumerate}
\end{exercise}

\begin{exercise}{Nontriviality of~$A^\sim$}%
Let~$A$ be a ring. In Proposition~\ref{prop:immediate-consequences}, we gave an
abstract proof that~$A^\sim$ is (local and hence) nontrivial in the sense that~$1
\neq 0$. This proof did not need to assume that~$A$ itself is nontrivial. Give a new proof of this
fact, by verifying directly~$1 \models \neg(1 =_{A^\sim} 0)$ using the
semantics given in Table~\ref{table:algebraic-kripke-joyal}.
\end{exercise}

\begin{exercise}{A generalization of the field property}%
Let~$A$ be a ring. We extend the sheaf semantics to infinitary first-order
formulas by adding the following clauses to
Table~\ref{table:algebraic-kripke-joyal}:
\begin{center}\tablefont
\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
  $f \models \bigwedge_{k \in K} \varphi_k$ &iff&
    for all indices~$k \in K$, $f \models \varphi_k$ \\
  $f \models \bigvee_{k \in K} \varphi_k$ &iff&
    there exists a partition~$f^n = fg_1 + \cdots + fg_m$ such that, \\
  &&\quad for each~$i$, there is an index~$k \in K$ such that~$fg_i \models \varphi_k$
\end{tabular}
\end{center}
Theorem~\ref{thm:basic-properties-sheaf-semantics-algebraic}
also apllies to this extension, and moreover, this extension in sound with
respect to infinitary intuitionistic logic.
\begin{enumerate}
\item Prove that:
\[ 1 \models \forall f\?A^\sim\_ \forall x\?A^\sim\_
  \bigl((\exists z\?A^\sim\_ fz = 1) \Rightarrow x = 0\bigr) \Longrightarrow \bigvee_{n \in \NN} (f^n x = 0). \]
\item Deduce that~$A^\sim$ is ``almost a field'' in the sense that any nonunit
is nilpotent.
\item Show that~$A$ is reduced iff~$A^\sim$ is.
\end{enumerate}
\end{exercise}
